Types from data: \textnormal{Making structured data first-class citizens in F\#}

### 2016-02-12 pldi-2016/paper.tex
Corrections

\begin{figure*}[t]
\noindent
\begin{equation*}
\begin{array}{rcll}
\ident{csh}(\sigma, \sigma) &=& \sigma & (\emph{eq})\\
\ident{csh}([\sigma_1], [\sigma_2]) &=& [\ident{csh}(\sigma_1, \sigma_2)] & (\emph{list}) \\
\ident{csh}(\bot, \sigma) = \ident{csh}(\sigma, \bot) &=& \sigma & (\emph{bot})\\
\ident{csh}(\kvd{null}, \sigma) = \ident{csh}(\sigma, \kvd{null}) &=& \addopt{\sigma}& (\emph{null}) \\
\ident{csh}(\kvd{any}, \sigma) = \ident{csh}(\sigma, \kvd{any}) &=& \kvd{any} & (\emph{top})\\
\ident{csh}(\kvd{float}, \kvd{int}) = \ident{csh}(\kvd{int}, \kvd{float}) &=& \kvd{float} & (\emph{num})\\
\ident{csh}(\sigma_2, \kvd{nullable}\langl \hat{\sigma_1} \rangl) = \ident{csh}(\kvd{nullable}\langl \hat{\sigma_1} \rangl, \sigma_2) &=& \addopt{\ident{csh}(\hat{\sigma_1}, \sigma_2)} & (\emph{opt})\\
\ident{csh}(\nu \; \{ \nu_1 \!:\! \sigma_1,  \ldots \;, \nu_n \!:\! \sigma_n \}, 
  \nu \; \{ \nu_1 \!:\! \sigma'_1, \; \ldots \;, \nu_n \!:\! \sigma'_n \}) &=& 
  \nu \; \{ \nu_1 \!:\! \ident{csh}(\sigma_1, \sigma'_1), \; \ldots \; , \nu_n \!:\! \ident{csh}(\sigma_n, \sigma'_n) \} & (\emph{recd}) \\
\ident{csh}(\sigma_1, \sigma_2) &=& \kvd{any} \quad(\textnormal{when}~\sigma_1\neq\nu\;\{ \ldots \}~\textnormal{or}~\sigma_2\neq\nu\;\{ \ldots \} ) & (\emph{any})\\
\end{array}
\end{equation*}
%
\begin{equation*}
\quad\qquad
\begin{array}{rcll}
 \addopt{\hat{\sigma}} &\narrow{=}& \kvd{nullable}\langl\hat{\sigma}\rangl &(\textnormal{non-nullable shapes})\\
 \addopt{\sigma} &\narrow{=}& \sigma &(\textnormal{otherwise})
\end{array}
\qquad\qquad
\begin{array}{rcll}
 \dropopt{\kvd{nullable}\langl\hat{\sigma}\rangl} &\narrow{=}& \hat{\sigma} &(\textnormal{nullable shape})\\
 \dropopt{\sigma} &\narrow{=}& \sigma &(\textnormal{otherwise})
\end{array}
\end{equation*}

@@

\noindent
The fact that the rules of \ident{csh} are matched from top to bottom resolves the ambiguity 
between certain rules. Most importantly (\emph{any}) is used only as the last resort.
When finding a common shape of two records (\emph{recd}) we find common preferred shapes of their 
respective fields. The \ident{csh} function is not defined for two records with mis-matching 
fields, however, the fields can always be made to match, through the shape inference 
from sample (\S\ref{sec:formal-inferval}), \emph{before} finding a common preferred shape 
using $\ident{csh}$.

We can find a common shape of two different numbers (\emph{num}); for two collections,
we combine their elements (\emph{list}). When one shape is nullable (\emph{opt}), we find
the common non-nullable shape and ensure the result is nullable using $\addopt{-}$, which
is also applied when one of the shapes is \kvd{null} (\emph{null}).

@@

\noindent
\begin{equation*}
\begin{array}{lcl}
  \\[-0.5em]
 d &\narrow{=}& i \lsep f \lsep s \lsep \kvd{true} \lsep \kvd{false} \lsep \kvd{null} \\[0.1em]
   &\narrow{|}& [d_1; \ldots; d_n] \lsep \nu~\{ \nu_1 \mapsto d_1, \ldots, \nu_n \mapsto d_n \}
\end{array}
\end{equation*}
%
The definition includes primitive values ($i$ for integers, $f$ for floats
and $s$ for strings) and \kvd{null}. A collection is written as a 
list of values in square brackets. A record starts with a name $\nu$, followed by a 
sequence of field assignments $\nu_i \mapsto d_i$.

@@

This is sufficient to equate the record field labels and satisfy the pre-conditions in rule 
(\emph{recd}) when multiple record samples are merged. In practice, $\theta$ is found via row 
variable unification \cite{rows-remy}. We omit the details here. No $\rho_i$ variables remain after 
inference as the substitution chosen is ground.

@@

\begin{remark}[Stability of inference]
Assume we have a set of samples $d_1, \ldots, d_n$, provided type based on the samples
$\tau_1, e_1, L_1 = \sem{\semalt{d_1, \ldots, d_n}}$ and some user code $e$ written using
the provided type, such that $L_1; x:\tau_1\vdash e : \tau$.

@@

\begin{itemize}
\item[--] Primitive shapes can become nullable (i), \ident{int} can become \ident{float} (iii)
  or can become a part of labelled top type (ii).
\item[--] Record shape fields can change shape (recursively) and record may become a part
  of labelled top type (ii).
\item[--] For list and nullable shape, the shape of the value may change (we apply the 
  transformations recursively).
\item[--] For any shape, the original code will continue to work (label is never removed).
\end{itemize}
\vspace{-1.5em}
\end{proof}
%
\noindent
Intuitively, the first transformation is needed when the new sample makes a type optional.
This happens when it contains a \kvd{null} value or a record that does not contain a field 
that all previous samples have. The second transformation is needed when a shape $\sigma$ 
becomes $\kvd{any}\langle\sigma, \ldots\rangle$ and the third one is needed when $\ident{int}$
becomes \ident{float}.

>>>>

\begin{figure*}[t]
\noindent
\begin{equation*}
\begin{array}{rcll}
\ident{csh}(\sigma, \sigma) &=& \sigma & (\emph{eq})\\
\ident{csh}([\sigma_1], [\sigma_2]) &=& [\ident{csh}(\sigma_1, \sigma_2)] & (\emph{list}) \\
\ident{csh}(\bot, \sigma) = \ident{csh}(\sigma, \bot) &=& \sigma & (\emph{bot})\\
\ident{csh}(\kvd{null}, \sigma) = \ident{csh}(\sigma, \kvd{null}) &=& \addopt{\sigma}& (\emph{null}) \\
\ident{csh}(\kvd{any}, \sigma) = \ident{csh}(\sigma, \kvd{any}) &=& \kvd{any} & (\emph{top})\\
\ident{csh}(\kvd{float}, \kvd{int}) = \ident{csh}(\kvd{int}, \kvd{float}) &=& \kvd{float} & (\emph{num})\\
\ident{csh}(\sigma_2, \kvd{nullable}\langl \hat{\sigma_1} \rangl) = \ident{csh}(\kvd{nullable}\langl \hat{\sigma_1} \rangl, \sigma_2) &=& \addopt{\ident{csh}(\hat{\sigma_1}, \sigma_2)} & (\emph{opt})\\
\ident{csh}(\nu \; \{ \nu_1 \!:\! \sigma_1,  \ldots \;, \nu_n \!:\! \sigma_n \}, 
  \nu \; \{ \nu_1 \!:\! \sigma'_1, \; \ldots \;, \nu_n \!:\! \sigma'_n \}) &=& 
  \nu \; \{ \nu_1 \!:\! \ident{csh}(\sigma_1, \sigma'_1), \; \ldots \; , \nu_n \!:\! \ident{csh}(\sigma_n, \sigma'_n) \} & (\emph{recd}) \\
\ident{csh}(\sigma_1, \sigma_2) &=& \kvd{any} \quad(\textnormal{when}~\sigma_1\neq\nu\;\{ \ldots \}~\textnormal{or}~\sigma_2\neq\nu\;\{ \ldots \} ) & (\emph{any})\\
\end{array}
\end{equation*}

\begin{equation*}
\quad\qquad
\begin{array}{rcll}
 \addopt{\hat{\sigma}} &\narrow{=}& \kvd{nullable}\langl\hat{\sigma}\rangl &(\textnormal{non-nullable shapes})\\
 \addopt{\sigma} &\narrow{=}& \sigma &(\textnormal{otherwise})
\end{array}
\qquad\qquad
\begin{array}{rcll}
 \dropopt{\kvd{nullable}\langl\hat{\sigma}\rangl} &\narrow{=}& \hat{\sigma} &(\textnormal{nullable shape})\\
 \dropopt{\sigma} &\narrow{=}& \sigma &(\textnormal{otherwise})
\end{array}
\end{equation*}

@@

\noindent
The fact that the rules of \ident{csh} are matched from top to bottom resolves the ambiguity 
between certain rules. Most importantly (\emph{any}) is used only as the last resort.

When finding a common shape of two records (\emph{recd}) we find common preferred shapes of their 
respective fields. We can find a common shape of two different numbers (\emph{num}); for two collections,
we combine their elements (\emph{list}). When one shape is nullable (\emph{opt}), we find
the common non-nullable shape and ensure the result is nullable using $\addopt{-}$, which
is also applied when one of the shapes is \kvd{null} (\emph{null}).

@@

\noindent
\begin{equation*}
\begin{array}{lcl}
  \\[-0.5em]
 d &\narrow{=}& i \lsep f \lsep s \lsep \kvd{true} \lsep \kvd{false} \lsep \kvd{null} \\[0.1em]
   &\narrow{|}& [d_1; \ldots; d_n] \lsep \nu~\{ \nu_1 \mapsto d_1, \ldots, \nu_n \mapsto d_n \}
\end{array}
\end{equation*}

~

\noindent
The definition includes primitive values ($i$ for integers, $f$ for floats
and $s$ for strings) and \kvd{null}. A collection is written as a 
list of values in square brackets. A record starts with a name $\nu$, followed by a 
sequence of field assignments $\nu_i \mapsto d_i$.

@@

This is sufficient to equate the record field labels and satisfy the pre-conditions in rule 
(\emph{recd}) when multiple record shapes are combined. The \ident{csh} function is not defined for 
two records with mis-matching fields, however, the fields can always be made to match, through 
a substitution for row variables. In practice, $\theta$ is found via row 
variable unification \cite{rows-remy}. We omit the details here. No $\rho_i$ variables remain after 
inference as the substitution chosen is ground.

@@

\begin{remark}[Stability of inference]
Assume we have a set of samples $d_1, \ldots, d_n$, a provided type based on the samples
$\tau_1, e_1, L_1 = \sem{\semalt{d_1, \ldots, d_n}}$ and some user code $e$ written using
the provided type, such that $L_1; x:\tau_1\vdash e : \tau$.

@@

\begin{itemize}
\item[--] Primitive shapes can become nullable (i), \ident{int} can become \ident{float} (iii)
  or can become a part of labelled top type (ii).
\item[--] Record shape fields can change shape (recursively) and record may become a part
  of labelled top type (ii).
\item[--] For list and nullable shapes, the shape of the value may change (we apply the 
  transformations recursively).
\item[--] For the any shape, the original code will continue to work (the label is never removed).
\end{itemize}
\vspace{-1.5em}
\end{proof}
%
\noindent
Intuitively, the first transformation is needed when the new sample makes a type optional.
This happens when it contains a \kvd{null} value or a record that does not contain a field 
that all previous samples have. The second transformation is needed when a shape $\sigma$ 
becomes $\kvd{any}\langle\sigma, \ldots\rangle$ and the third one is needed when $\ident{int}$
becomes \ident{float}.
